---
title: "JSC370 Final Project"
date: "April 30, 2025"
author: "Matteo Guzzi"
output:
  html_document:
    df_print: paged
---

## Introduction
Toronto, being one of Canada's biggest cities, is faced with efficient urban mobility, safety, and infrastructural planning. Bicycle thefts are a persistent issue that affects cyclists across the city. As cycling grows in popularity as a sustainable mode of transportation, understanding the spatial patterns and socioeconomic factors associated with bike theft becomes increasingly important for urban planning and public safety initiatives. This study examines the relationship between neighborhood characteristics and bicycle theft rates using open data from the City of Toronto.

The analysis integrates three key datasets: Bicycle Thefts, Neighbourhood Profiles and Bicycle Shops. These datasets collectively provide a comprehensive view of Toronto's transportation ecosystem and its relationship to bicycle crime.

The bicycle thefts dataset consists of incident reports of thefts with the details of the location where thefts took place, the timing, and make, model, and price of the bikes. The spatial distribution of these thefts is valuable to determine danger zones and inform law enforcement agencies and cyclists about areas of vulnerability. As well, the Bicycle Shops dataset contains information on registered bicycle shops and repair shops, providing a picture of the availability of bike-related shops throughout the city. These shops serve not only the cycling public but also the local economy and cycling availability in different neighbourhoods.

The Neighbourhood Profiles dataset provides a socio-economic context to the analysis, including demographic, income, and housing data for 158 Toronto neighbourhoods. Initially created to assist government and community organizations with local planning, this data set allows further investigation into the variation of social and economic status between the city. Together with the four data sets, a more synthesized image can be established of Toronto's transportation setting, for instance, how traffic volume, theft danger, and bicycle infrastructure exist alongside broader neighbourhood characteristics.

Thus the research question formulated is; *what kind of socioeconomic indicators are there in areas with high bike theft? How do bike shops impact bike theft?*

I hypothesize that bicycle theft rates will be highest in neighborhoods with lesser population density and lower concentrations of bicycle shops, as these factors may provide both more targets and greater anonymity for thieves. Additionally, I expect that socioeconomic factors like income levels and housing types will show interesting correlations with theft patterns.


## Methods
All data were acquired from Toronto's Open Data Portal using R's $opendatatoronto$ and $httr$ packages. The datasets were retrieved using the city's API, which gives real-time data access in CSV format. The API was utilized to fetch the most current records with some automating data wrangling. 

After the data were obtained, they passed through a strict cleaning process Every dataset had its own structures and needed to undergo different transformations. 

The bicycle theft dataset was filtered to include only incidents from 2014 onward with reported bike values exceeding $50 to ensure data quality. Spatial coordinates were extracted and converted to an SF (simple features) object for geographic analysis. All spatial data (for all datasets) was projected using WGS84 (EPSG:4326) for consistency.

The Bicycle Shop dataset had store names, addresses, and geospatial coordinates. Coordinates were pulled and cleaned using tidyverse functions to be utilized in spatial joins.

The Neighbourhood Profiles dataset was by far the most difficult dataset. After multiple attempts I transposed so that each row represents a neighborhood, each with more than 1300 variables. Using $tidyr$ functions, I thus transposed the data, cleaned it, and converted percentage values to numeric format. Neighborhood boundaries were joined with profile data using official neighborhood codes. 



```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(tidytext)
library(tidyverse)
library(wordcloud2)
library(tm)
library(topicmodels)
library(knitr)
library(httr)
library(jsonlite)
library(tidyr)
library(dplyr)
library(leaflet)
library(readr)
library(kableExtra)
library(sf)
library(opendatatoronto)
library(plotly)
library(randomForest)
library(mgcv)

```


use the API to extrat 2 csv: traffic and neighborhood (socio-economic data)
```{r download data, message=FALSE, echo=FALSE, warning=FALSE}
#sources:
# https://open.toronto.ca/dataset/traffic-volumes-midblock-vehicle-speed-volume-and-classification-counts/ 
# https://open.toronto.ca/dataset/neighbourhood-profiles/ 
# https://open.toronto.ca/dataset/bicycle-thefts/ 
# https://open.toronto.ca/dataset/bicycle-shops/ 
# additional data (potential):
# https://open.toronto.ca/dataset/neighbourhoods/ boundaries to overlap 
# https://open.toronto.ca/dataset/cycling-network/ 


base_url <- "https://ckan0.cf.opendata.inter.prod-toronto.ca"
package_url <- paste0(base_url, "/api/3/action/package_show")

fetch_data2 <- function(dataset_id) {
  response <- GET(package_url, query = list(id = dataset_id))
  metadata <- fromJSON(content(response, "text", encoding = "UTF-8"))
  resources <- metadata$result$resources
  datastore_resource <- resources %>%
    filter(datastore_active == TRUE) %>%
    select(id) %>%
    slice(1) %>%
    pull()
  
  csv_url <- paste0(base_url, "/datastore/dump/", datastore_resource)
  
  read_csv(csv_url)
}
fetch_data <- function(dataset_id) {
  response <- GET(package_url, query = list(id = dataset_id))
  metadata <- fromJSON(content(response, "text", encoding = "UTF-8"))
  resources <- metadata$result$resources
  csv_resource <- resources[resources$format == "CSV" & resources$url_type == "datastore", ]
  csv_url <- csv_resource$url[1]
  read_csv(csv_url)
}


neighbourhood_data <- fetch_data("neighbourhood-profiles")
traffic_data <- fetch_data("traffic-volumes-midblock-vehicle-speed-volume-and-classification-counts")
thefts_data <- fetch_data2("bicycle-thefts")
shops_data <- fetch_data2("bicycle-shops")

package <- show_package("neighbourhoods")
resources <- list_package_resources("neighbourhoods")
boundaries_data = get_resource(resources$id[1])
```



```{r feedback data, include=FALSE}
library(opendatatoronto)
resources <- list_package_resources("945444df-c9cf-4a9d-97a8-ee8b0feeb419")

data <- get_resource(resources$id[1])

# cannot do text wrangling - feedback is multiple 


```



```{r data summary, eval=FALSE, echo=FALSE, warning=FALSE}
#check for sus values
head(neighbourhood_data)
head(traffic_data)
head(thefts_data)
head(shops_data)


summary(neighbourhood_data)
summary(thefts_data)

# some bikes cost too little 
# first date is 1975?

summary(shops_data)

unique(neighbourhood_data$Topic)
colnames(neighbourhood_data)
unique(neighbourhood_data$Topic)
```




```{r data cleaning, echo=FALSE, warning=FALSE}
# clean data
thefts <- thefts_data %>%
  drop_na() %>%
  rename(
    offense = PRIMARY_OFFENCE,
    date = OCC_DATE,
    year = OCC_YEAR,
    make = BIKE_MAKE,
    model = BIKE_MODEL,
    type = BIKE_TYPE,
    speed = BIKE_SPEED,
    color = BIKE_COLOUR,
    cost = BIKE_COST,
    longitude = LONG_WGS84,
    latitude = LAT_WGS84
  ) %>%
  filter(cost >= 50,
         year > 2013) %>%  
  mutate(entity = "Theft")


extract_coordinates <- function(geometry) {
  coords <- fromJSON(geometry)$coordinates
  return(data.frame(longitude = coords[1], latitude = coords[2]))
}

shops <- shops_data %>%
  rename(
    shop_id = `_id`,
    shop_name = NAME,
    address = ADDRESS_FULL,
    postal_code = POSTAL_CODE,
    city = CITY
  ) %>%
  mutate(entity = "Shop") %>% 
  mutate(coords = map(geometry, ~ fromJSON(.x)$coordinates)) %>% 
  mutate(
    longitude = map_dbl(coords, 1),  
    latitude = map_dbl(coords, 2)   
  ) %>%
  select(-geometry, -coords)

#convert to spatial dataframe  


thefts_sf <- st_as_sf(thefts, coords = c("longitude", "latitude"), crs = 4326)
shops_sf <- st_as_sf(shops, coords = c("longitude", "latitude"), crs = 4326)



```

```{r neighborhood data cleaning and joining,echo = FALSE, warning=FALSE}
# Clean and transpose neighborhood data
neighborhood_clean <- neighbourhood_data %>%
  select(-c(1:4)) %>%
  pivot_longer(cols = -Characteristic, names_to = "Neighborhood", values_to = "Value") %>%
  pivot_wider(names_from = "Characteristic", values_from = Value) %>%
  mutate(across(
    .cols = -c("Neighborhood", "TSNS2020 Designation"),
    .fns = ~ readr::parse_number(as.character(.x))
  )) %>%
  rename("Population Change % 2011-2016" = "Population Change 2011-2016",
         "Neighborhood Number" = "Neighbourhood Number",
         "Pop2016" = "Population, 2016")
  
# Get neighborhood boundaries and join with profile data
neighborhoods_sf <- st_as_sf(boundaries_data) %>%
  st_transform(4326)

neighborhood_full <- neighborhoods_sf %>%
  mutate(across(
    .cols = c("AREA_LONG_CODE"),
    .fns = ~ readr::parse_number(as.character(.x))
  )) %>%
  left_join(neighborhood_clean, by = c("AREA_LONG_CODE" = "Neighborhood Number")) %>%
  select(-c(AREA_ID, AREA_ATTR_ID, PARENT_AREA_ID, AREA_SHORT_CODE, 
            AREA_NAME, AREA_DESC, CLASSIFICATION, 
            OBJECTID, "TSNS2020 Designation")) %>%
  filter(AREA_LONG_CODE <= 140) %>%
  janitor::clean_names()

```

### Exploratory Analysis
Initial exploration involved creating summary statistics table and visualizations. This initially included some simpler plots such as distribution plots of bike theft costs and temporal patterns and leaflet maps showing spatial distributions of thefts and bike shops. Bike thefts had to be logged due to the daa being ver left skewed (ie only a few bikes cost very much). 

Key tools included ggplot2 for static visualizations, plotly for interactive plots, and leaflet for spatial mapping. Summary statistics were presented using kableExtra for publication-quality tables.

After the data was merged saptially a neighborhood-level theft rates (thefts per 1000 residents) visualziation was created as well as a correlation analysis between theft rates and neighborhood characteristics.

### Modeling Approach
Before model fitting the variables in the data had to bed reduced. Using a heatmap and filtering for the most significant variables the columns were reduced for more than 1300 to just a selected few. 

Afterwards, three different modeling techniques were employed to predict high-theft neighborhoods (defined as those above the median theft rate):

Logistic Regression: A generalized linear model using binomial family with logit link function. This provides interpretable coefficients showing how each predictor affects the log-odds of being a high-theft neighborhood.

Random Forest: An ensemble method combining multiple decision trees, with 500 trees grown using random subsets of predictors. This non-parametric approach handles complex interactions and provides variable importance measures.

Generalized Additive Model (GAM): A semi-parametric extension of GLMs that allows for smooth, non-linear relationships between predictors and outcome using spline functions (k=3 basis dimensions).

Models were trained using a random 70% split of neighborhoods, with the remaining 30% reserved as a holdout test set. Prior to modeling, preprocessing steps were applied: all numeric predictors were standardized by subtracting the mean and dividing by the standard deviation. Missing values in the predictors were imputed using the mean calculated from the training set to avoid data leakage. Categorical variables were converted to numeric form, and variables containing only missing values were removed.

Performance was assessed using classification accuracy and confusion matrices. For the GAM, we examined the significance of smooth terms and the proportion of deviance explained. The random forest provided variable importance scores indicating each predictor's contribution to model accuracy.


## Results 
The cleaned dataset included 124 out of 158 neighborhoods with complete data. On average, the cost of a stolen bicycle was approximately $1000, with a median cost of $700. Theft rates varied substantially across neighborhoods, ranging from 0.03 to a whopping 129.49 thefts per 1,000 residents. The highest rates of bike theft were concentrated in downtown areas (university, chinatown, and moss park were the top-3). Spatial analysis revealed that bike shops showed moderate spatial correlation with theft locations, suggesting potential hotspots.

Interactive maps demonstrated some clustering of theft incidents in central business districts and near major transit hubs. Temporal analysis indicated consistent theft patterns across years, with slight seasonal variation, particularly higher rates during the warmer months.

The logistic regression model achieved 89.2% accuracy on the test data, with an RMSE of 0.3247 and MAE of 0.1067, outperforming the generalized additive model (GAM), which achieved 73.0% accuracy, an RMSE of 0.4275, and MAE of 0.3274. The random forest model provided additional insight into important predictors of bike theft risk. The variable importance plot indicates that single young adults living at home and bicycle commuting rates were among the strongest predictors. Other highly important predictors included employment rate, average household size, and percentage of persons living alone. Notably, population density and walking rates also ranked among the top predictors, emphasizing the influence of urban living environments and mobility patterns on theft risk.

The GAM results further supported the importance of urban density. The highly significant smooth term for population density (p < 0.0001) confirmed a strong non-linear relationship between density and theft risk. However, the model explained only 20.8% of deviance, suggesting that other unmeasured factors also contribute to variations in theft patterns.

```{r summary stats, message=FALSE, echo=FALSE, warning=FALSE}
## Summary statistics in tables


num_neighbourhoods <- ncol(neighbourhood_data) - 4
overall_avg_population <- neighbourhood_data %>%
  filter(Category == "Population",
         Topic == "Population and dwellings",
         Characteristic == "Population, 2016") %>%
  select(-c(1:6)) %>%  
    mutate(across(everything(), ~ as.numeric(gsub("[^0-9]", "", .)))) %>%
  unlist() %>%
  mean(na.rm = TRUE)

neighbourhood_summary <- tibble(
  `Number of Neighbourhoods` = num_neighbourhoods,
  `Overall Avg Population` = round(overall_avg_population, 0)
)


thefts_summary <- thefts_sf %>%
  summarise(
    Total_Thefts = n(),
    Avg_Bike_Cost = mean(cost, na.rm = TRUE),
    Min_Bike_Cost = min(cost, na.rm = TRUE),
    Max_Bike_Cost = max(cost, na.rm = TRUE),
    Median_Bike_Cost = median(cost, na.rm = TRUE),
    Variance_Bike_Cost = var(cost, na.rm = TRUE),
    Q1_Bike_Cost = quantile(cost, 0.25, na.rm = TRUE),
    Q3_Bike_Cost = quantile(cost, 0.75, na.rm = TRUE)
  )

shops_summary <- shops_sf %>%
  summarise(
    Total_Shops = n(),
    Unique_Postal_Codes = n_distinct(postal_code),
    Most_Common_Postal_Code = names(sort(table(postal_code), decreasing = TRUE)[1])
  )



# Thefts Data Summary
kable(thefts_summary, caption = "Table 1: Summary of Bicycle Thefts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"), full_width = FALSE) %>%
  footnote(general = "This table summarizes the cost distribution of stolen bikes.")

# Shops Data Summary
kable(shops_summary, caption = "Table 2: Summary of Bicycle Shops in Toronto") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"), full_width = FALSE) %>%
  footnote(general = "This table displays the number of bike shops and unique postal codes covered.")

# Neighbourhood Summary
kable(neighbourhood_summary, caption = "Table 3: Neighbourhood Statistics (Count & Average Population)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"), full_width = FALSE) %>%
  footnote(general = "This table summarizes the number of neighbourhoods and their average population.")


```
 
 
```{r basic maps, eval = FALSE, message=FALSE, echo=FALSE, warning=FALSE}
## Basic maps


leaflet(thefts_sf) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = 4,
    color = "red",
    fillOpacity = 0.7,
    popup = ~paste("Bike Model:", model, "<br>Bike Cost: $", cost)
  ) %>%
  addLegend(position = "bottomright", colors = "red", labels = "Bike Thefts", title = "Bike Thefts Data")



leaflet(shops_sf) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = 4,
    color = "green",
    fillOpacity = 0.7,
    popup = ~paste("Shop Name:", shop_name, "<br>Address:", address)
  ) %>%
  addLegend(position = "bottomright", colors = "green", labels = "Bike Shops", title = "Bike Shops Data")


```
 
 
 
```{r basic plots, message=FALSE, echo=FALSE, warning=FALSE}

# Bike Theft Costs Histogram (Interactive)
thefts_sf_log <- thefts_sf %>%
  mutate(log_cost = log1p(cost))  # log(1 + cost) to avoid log(0) issues

p2 <- ggplot(thefts_sf_log, aes(x = log_cost)) +
  geom_histogram(bins = 30, fill = "red", color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Bike Theft Costs (Log Scale)",
    x = "Log(Bike Cost + 1)",
    y = "Frequency"
  )

ggplotly(p2, tooltip = "text") %>%
  layout(hoverlabel = list(bgcolor = "white"))


# Bike Thefts Over Years Bar Plot (Interactive)
p3 <- ggplot(thefts_sf, aes(x = year)) +
  geom_bar(fill = "red", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Bike Thefts Over the Years",
    x = "Year",
    y = "Number of Thefts"
  )

ggplotly(p3, tooltip = "count") %>% 
  layout(hoverlabel = list(bgcolor = "white"))

```



```{r data merging, message=FALSE, echo=FALSE, warning=FALSE}

# Merge datasets using spatial join (nearest feature)

thefts_clean <- thefts_sf %>%
  mutate(entity = "Theft")

shops_clean <- shops_sf %>%
  mutate(entity = "Shop")

##### full join
merged_sf <- bind_rows(thefts_clean, shops_clean)



```


```{r plot maps v2, message=FALSE, echo=FALSE, warning=FALSE}
leaflet(merged_sf) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = 4,
    color = ~ case_when(
      entity == "Theft" ~ "red",
      entity == "Shop" ~ "green"
    ),
    popup = ~entity
  ) %>%
  addLegend(position = "bottomright",
            colors = c("red", "green"),
            labels = c("Theft", "Shop"),
            title = "Entity Type")


```

```{r bike thefts by neighborhoods, message=FALSE, echo=FALSE, warning=FALSE}
thefts_with_neighborhood <- thefts_sf %>%
  st_join(neighborhood_full, join = st_intersects, left = FALSE) %>%
  janitor::clean_names()
  

# Calculate theft density per neighborhood
theft_density <- thefts_with_neighborhood %>%
  st_drop_geometry() %>%
  count(neighborhood, name = "theft_count") %>%
  left_join(
    neighborhood_full,
    by = "neighborhood"
  ) %>%
  mutate(
    theft_rate = theft_count / pop2016,
    theft_rate_per_1000 = theft_rate * 1000  
  ) 

top_20_theft <- theft_density %>%
  arrange(desc(theft_rate)) %>% 
  slice(1:20)                   

p_bar <- ggplot(top_20_theft, aes(x = reorder(neighborhood, theft_rate_per_1000), y = theft_rate_per_1000)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 20 Neighborhoods by Bike Theft Rate",
       x = "Neighborhood",
       y = "Thefts per 1000 people") +
  theme_minimal()

ggplotly(p_bar)

bottom_20_theft <- theft_density %>%
  arrange(theft_rate) %>%  
  slice(1:20)           


ggplot(bottom_20_theft, aes(x = reorder(neighborhood, -theft_rate_per_1000), y = theft_rate_per_1000)) +
  geom_col(fill = "darkgreen") + 
  coord_flip() +
  labs(title = "20 Safest Neighborhoods for Bike Theft",
       x = "Neighborhood",
       y = "Thefts per 1000 people") +
  theme_minimal()

## also do shops/neighborhood and create a map


```


```{r shops & bikes by neighborhoods with map, message=FALSE, echo=FALSE, warning=FALSE}
shops_with_neighborhood <- shops_sf %>%
  st_join(neighborhood_full, join = st_intersects, left = FALSE) %>%
  janitor::clean_names()

# Calculate number of shops per neighborhood
shop_density <- shops_with_neighborhood %>%
  st_drop_geometry() %>%
  count(neighborhood, name = "shop_count") %>%
  left_join(
    neighborhood_full,
    by = "neighborhood"
  ) %>%
  mutate(
    shops_per_1000 = shop_count / pop2016 * 1000
  )


# Merge shop density into spatial neighborhood data
neighborhood_with_shops <- neighborhood_full %>%
  left_join(shop_density %>% select(neighborhood, shops_per_1000), by = "neighborhood")



```

```{r spatial maps joined data, message=FALSE, echo=FALSE, warning=FALSE}
# Plot shops per neighborhood per 1000ppl
ggplot(neighborhood_with_shops) +
  geom_sf(aes(fill = shops_per_1000)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey80") +
  labs(title = "Bike Shops per 1000 Residents by Neighborhood",
       fill = "Shops per 1000") +
  theme_minimal()

# Plot thefts per neighborhood per 1000ppl
neighborhood_with_thefts <- neighborhood_full %>%
  left_join(theft_density %>% select(neighborhood, theft_rate_per_1000), by = "neighborhood")


p1 <- ggplot(neighborhood_with_thefts) +
  geom_sf(aes(fill = theft_rate_per_1000)) +
  scale_fill_viridis_c(option = "magma", na.value = "grey80") +
  labs(title = "Bike Thefts per 1000 Residents by Neighborhood",
       fill = "Thefts per 1000") +
  theme_minimal()

ggplotly(p1)

## plot joined data
combined_data <- neighborhood_full %>%
  left_join(theft_density %>% select(neighborhood, theft_rate_per_1000), by = "neighborhood") %>%
  left_join(shop_density %>% select(neighborhood, shops_per_1000), by = "neighborhood") %>%
  mutate(pop2016 = as.numeric(pop2016))

#bike shops vs. thefts, size = population
p <- ggplot(combined_data, aes(
    x = shops_per_1000,
    y = theft_rate_per_1000,
    size = pop2016,
    text = paste0(
      "<b>Neighborhood:</b> ", neighborhood, "<br>",
      "<b>Bike Shops per 1000:</b> ", round(shops_per_1000, 2), "<br>",
      "<b>Thefts per 1000:</b> ", round(theft_rate_per_1000, 2), "<br>",
      "<b>Population:</b> ", pop2016
    )
  )) +
  geom_point(alpha = 0.7, color = "steelblue") +
  scale_size_continuous(range = c(1, 10), name = "Population (2016)") +
  labs(
    title = "Relationship Between Bike Shops, Bike Thefts, and Population by Neighborhood",
    x = "Bike Shops per 1000 Residents",
    y = "Bike Thefts per 1000 Residents"
  ) +
  theme_minimal()

# Turn it into an interactive plotly plot
ggplotly(p, tooltip = "text") %>%
  layout(
    hoverlabel = list(bgcolor = "white"),
    title = list(
      text = "<b>Relationship Between Bike Shops, Bike Thefts, and Population by Neighborhood</b>"
    )
  )


```





```{r feature selection, message=FALSE, echo=FALSE, warning=FALSE}
#find out what features best predict bike theft by neighbourhood

#Correlation analysis for numeric variables
numeric_theft_density <- theft_density %>%
  st_drop_geometry() %>%
  select(where(is.numeric))%>%
  mutate(across(everything(), ~ {
    x <- as.numeric(.)
    ifelse(is.infinite(x) | is.na(x), mean(x, na.rm = TRUE), x)
  })) %>%
  select_if(~ !any(is.na(.)))

cor_matrix <- as.matrix(cor(numeric_theft_density))
target_cor <- cor_matrix[,"theft_rate"]
strong_vars <- names(which(abs(target_cor) > 0.6))

heatmap(cor_matrix[strong_vars, strong_vars], Colv = NA, Rowv = NA)


#select additional potential variables
demo_features <- c("pop2016", "population_density_per_square_kilometre",
                   "youth_15_24_years", "working_age_25_54_years","average_household_size",
                   "persons_living_alone_per_cent")
econ_features <- c("median_total_income_in_2015_among_recipients",
                   "in_low_income_based_on_the_low_income_measure_after_tax_lim_at","employment_rate")
housing_features <- c("apartment_in_a_building_that_has_five_or_more_storeys","row_house", "condominium",
                      "average_weeks_worked_in_reference_year")
transport_features <- c("bicycle", "public_transit", "no_fixed_workplace_address",
                        "commute_within_census_subdivision_csd_of_residence")
time_features <- c("worked_full_year_full_time", "between_5_a_m_and_5_59_a_m", "between_12_p_m_and_4_59_a_m")

selected_data <- theft_density %>%
  select(
    neighborhood, all_of(demo_features), all_of(econ_features),
    all_of(housing_features), all_of(transport_features),
    all_of(time_features),
    all_of(strong_vars)
  ) %>%
  # Convert percentages to numeric
  mutate(across(ends_with("_percent"), ~ as.numeric(gsub("%", "", .))/100))


```

```{r split data & preprocess, message=FALSE, echo=FALSE, warning=FALSE}
set.seed(5858)

# add binary classification (high vs low theft rate)
selected_data <- selected_data %>%
  mutate(high_theft = as.numeric(theft_rate > median(theft_rate))) %>%
  select(-theft_rate) %>%
  st_drop_geometry()

#split
train <- sample(1:nrow(selected_data), round(0.7 * nrow(selected_data)))

train_data <- selected_data[train, ]
test_data <- selected_data[-train, ]

#preprocess
preprocess_data <- function(df, train_mean = NULL) {
  df_processed <- df %>%
    select(-neighborhood) %>%
    mutate(across(where(is.character), as.factor)) %>%
    mutate(across(where(is.factor), as.numeric)) %>%
    mutate(across(where(is.numeric) & !high_theft, ~ {
      x <- .x
      if(is.null(train_mean)) {
        x_mean <- mean(x, na.rm = TRUE)
      } else {
        x_mean <- train_mean
      }
      x[is.na(x)] <- x_mean
      scale(x, center = mean(x, na.rm = TRUE), scale = sd(x, na.rm = TRUE))
    }))
  
  # Remove columns with all NA values
  df_processed <- df_processed[, colSums(is.na(df_processed)) < nrow(df_processed)]
  
  return(df_processed)
}


train_processed <- preprocess_data(train_data)
test_processed <- preprocess_data(test_data)

```




```{r logistic regression model, message=FALSE, echo=FALSE, warning=FALSE}
# Binary outcome: high-value thefts (above median)


logit_model <- glm(high_theft ~ ., 
                  data = train_processed, 
                  family = binomial(link = "logit"),
                  control = list(maxit = 100))

```


```{r random forest, echo = FALSE, warning=FALSE, message=FALSE}


rf_model <- randomForest(x = train_processed[, -which(names(train_processed) == "high_theft")],
                        y = as.factor(train_processed$high_theft),
                        ntree = 500, 
                        importance = TRUE)

```

```{r GAM model, echo = FALSE, warning=FALSE, message=FALSE}


train_processed$income

gam_model <- gam(high_theft ~ 
                 s(population_density_per_square_kilometre, k = 3) + 
                 s(public_transit, k = 3),
               data = train_processed,
               family = binomial(),
               method = "REML")
```


```{r Model Evaluation, echo = FALSE}

# Model Evaluation 
evaluate_model <- function(model, test_data) {
  # Get predicted probabilities
  preds_prob <- predict(model, newdata = test_data, type = "response")
  
  # Convert to class predictions (0/1)
  preds_class <- ifelse(preds_prob > 0.5, 1, 0)
  
  # Calculate accuracy
  accuracy <- mean(preds_class == test_data$high_theft)
  
  # Calculate RMSE and MAE
  rmse <- sqrt(mean((preds_prob - test_data$high_theft)^2))
  mae <- mean(abs(preds_prob - test_data$high_theft))
  
  # Create confusion matrix manually
  confusion <- table(Predicted = preds_class, Actual = test_data$high_theft)
  
  return(list(
    accuracy = accuracy,
    rmse = rmse,
    mae = mae,
    confusion = confusion
  ))
}



# Compare performance
logit_perf <- evaluate_model(logit_model, test_processed)
gam_perf <- evaluate_model(gam_model, test_processed)

model_results <- data.frame(
  Model = c("Logistic Regression", "GAM"),
  Accuracy = c(logit_perf$accuracy, gam_perf$accuracy),
  RMSE = c(logit_perf$rmse, gam_perf$rmse),
  MAE = c(logit_perf$mae, gam_perf$mae)
)

# Display the results in a kable
kable(model_results, 
      col.names = c("Model", "Accuracy", "RMSE", "MAE"), 
      caption = "Table 4: Model Evaluation Results")


#Random Forest Accuracy
rf_preds <- predict(rf_model, newdata = test_processed, type = "response")
rf_preds_class <- ifelse(rf_preds > 0.5, 1, 0)

rf_accuracy <- mean(rf_preds_class == test_processed$high_theft)
rf_confusion <- table(Predicted = rf_preds_class, Actual = test_processed$high_theft)

cat("Random Forest Accuracy:", rf_accuracy, "\n")


# Variable importance
importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

importance_df <- importance_df %>%
  arrange(desc(MeanDecreaseGini)) %>%
  slice(3:n())

ggplot(importance_df, aes(x = reorder(Variable, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +  # Flip axes for easier reading
  theme_minimal(base_size = 14) +
  labs(
    title = "Random Forest Feature Importance",
    x = "Feature",
    y = "Mean Decrease Gini"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10)
  )


summary(gam_model)  
```





## Conclusions
This analysis identified population density as the strongest predictor of bicycle theft rates in Toronto neighborhoods, with denser urban areas exhibiting significantly higher theft risks. These findings align with routine activity theory in criminology, which posits that crime occurs when motivated offenders, suitable targets, and a lack of guardianship converge—conditions more likely in high-density environments. The concentration of bike thefts in central areas of Toronto, particularly around the university, Chinatown, and Moss Park, further supports this theory.

In addition to population density, the analysis revealed that public transit infrastructure plays a secondary role in influencing theft rates. The proximity of transit hubs may facilitate theft by offering easy access for thieves to transport stolen bicycles, or it may increase the concentration of potential targets in these areas. However, the presence of bike shops showed only weak correlation with theft incidents, contrary to initial expectations, indicating that other factors may be more influential in determining theft patterns.

The logistic regression model, which achieved an accuracy of 89.2%, outperformed the generalized additive model (GAM), which achieved an accuracy of 73.0%. The logistic regression model also showed a lower root mean square error (RMSE) and mean absolute error (MAE), suggesting a better fit for the data. The random forest model helped identify key predictors of bike theft risk, such as young adults living at home, bicycle commuting rates, and urban density. These findings highlight the influence of both socio-economic factors and the built environment on theft risk.

Key limitations of the study include the reliance on reported theft data, which may be subject to underreporting, and the lack of data on bike parking infrastructure, which could be a significant factor in preventing theft. Additionally, the temporal resolution of the data was limited to annual patterns, which may overlook seasonal fluctuations in theft activity. There may also be omitted variables, such as policing presence, that were not captured in the analysis.

Based on these findings, several policy implications emerge. Targeted theft prevention strategies in high-density neighborhoods, where theft rates are highest, are essential. Additionally, the implementation of secure bike parking near major transit hubs could mitigate theft risks. Public awareness campaigns in vulnerable areas, particularly those with high theft rates, could also help reduce incidents.

Future research could benefit from incorporating higher-resolution temporal data to capture more granular patterns in theft activity. Further analysis of bike lane infrastructure metrics and their relationship with theft rates could provide valuable insights. Additionally, studying recovery rates and conducting victimization surveys would address reporting biases and enhance the accuracy of theft data.

Overall, this study demonstrates how open data and spatial analysis can inform urban safety planning. By providing evidence-based insights, it offers valuable recommendations for reducing bicycle theft in Toronto and improving the safety of cyclists in the city.


*link to repo*: https://github.com/guzzim2022/jsc370-guzzi-finalproject 





