---
title: "JSC370 Final Project"
author: "Matteo Guzzi"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

You can view my full report with figures and analysis [here](final-proj-jsc370.pdf).


```{r source file, echo = FALSE, message=FALSE, warning=FALSE, include = FALSE}
library(plotly)
library(knitr)

knitr::purl("final-proj-jsc370.Rmd", output = "analysis_script.R", documentation = 0, quiet = TRUE) # convert rmd to R
source("analysis_script.R")

```



## Project Description

### Problem Statement

Toronto, being one of Canada's largest cities, faces challenges in urban mobility, safety, and infrastructural planning. Bicycle theft has become a significant issue, affecting cyclists across the city. As cycling continues to grow in popularity, it is crucial to understand the factors driving bicycle theft to help improve safety for cyclists.

This project investigates how socio-economic indicators and the presence of bicycle shops influence bicycle theft rates. The goal is to answer the following research questions:

1. What socio-economic indicators are present in neighborhoods with high bicycle theft rates?
2. How does the presence of bicycle shops impact bicycle theft?

The hypothesis is that neighborhoods with lower population density and fewer bicycle shops may have higher theft rates due to increased anonymity for thieves and a lack of nearby bike resources.

### Data Overview

The analysis uses the following datasets:

1. Bicycle Thefts Dataset: Incident reports of stolen bicycles, including the location, timing, and details about the bikes (make, model, value).
2. Bicycle Shops Dataset: Information about registered bicycle shops and repair shops across Toronto.
3. Neighbourhood Profiles Dataset: Socio-economic data for 158 neighborhoods in Toronto, covering demographics, income levels, and housing types.


The final analysis was performed using R packages tools such as `tidyverse`, `sf`, and `ggplot2` for data visualization and spatial analysis.



## Interactive Visualizations


### Top 20 Neighborhoods by Theft Rate
```{r plot 1, echo=FALSE}
p_bar <- ggplot(top_20_theft, aes(x = reorder(neighborhood, theft_rate_per_1000), y = theft_rate_per_1000)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = "Neighborhood", y = "Thefts per 1000 people")
ggplotly(p_bar)
```

### Bike Thefts per 1000 Residents (Map)
```{r plot 2, echo=FALSE}
# From 'spatial maps joined data' chunk
p1 <- ggplot(neighborhood_with_thefts) +
  geom_sf(aes(fill = theft_rate_per_1000)) +
  scale_fill_viridis_c(option = "magma") +
  labs(fill = "Thefts per 1000")
ggplotly(p1)
```

